---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2024"
subtitle: "Assignment 3 - Due date 02/01/24"
author: "David Robinson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
```{r}

```
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp24.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2022 **Monthly** Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. 

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here

library(forecast)
library(tseries)
library(dplyr)
library(readxl)
library(ggplot2)
library(Kendall)
library(cowplot)

```

```{r}
#Importing data set

getwd()

raw_energy_data <- read_excel(path="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",col_names=FALSE,skip=12)

colnames(raw_energy_data)=c("Month",
                            "Wood Energy Production",
                            "Biofuels Production",
                            "Total Biomass Energy Production",
                            "Total Renewable Energy Production",
                            "Hydroelectric Power Consumption",
                            "Geothermal Energy Consumption",
                            "Solar Energy Consumption",
                            "Wind Energy Consumption",
                            "Wood Energy Consumption",
                            "Waste Energy Consumption",
                            "Biofuels Consumption",
                            "Total Biomass Energy Consumption",
                            "Total Renewable Energy Consumption")

raw_energy_data <- raw_energy_data[,1:6]
raw_energy_data_dates <- raw_energy_data[,1]
raw_energy_data_others <- raw_energy_data[,4:6]
raw_energy_data <- cbind(raw_energy_data_dates,raw_energy_data_others)

head(raw_energy_data)

```


##Trend Component

### Q1
For each time series, i.e., Renewable Energy Production and Hydroelectric Consumption create three plots: one with time series, one with the ACF and with the PACF. You may use the some code form A2, but I want all the three plots side by side as in a grid.
(Hint: use function `plot_grid()` from the `cowplot` package)

```{r}

#Converting to time series
ts_energy_data <- ts(raw_energy_data[,2:4], start=c(1973,1), frequency=12)

head(ts_energy_data)
tail(ts_energy_data)

#Preparing Renewable Energy plots
renewable_ts_plot <- autoplot(ts_energy_data[,2]) +
  ggtitle("Renewable Energy Time Series") +
  xlab("Time") +
  ylab("Trillion Btu")

renewable_acf <- Acf(ts_energy_data[,2],lag.max=40,type="correlation",
                    plot=TRUE)

renewable_pacf <- Pacf(ts_energy_data[,2],lag.max=40,plot=TRUE)

#Preparing Hydroelectric Energy plots
hydroelectric_ts_plot <- autoplot(ts_energy_data[,3]) +
  ggtitle("Hydroelectric Energy Time Series") +
  xlab("Time") +
  ylab("Trillion Btu")

hydroelectric_acf <- Acf(ts_energy_data[,3],lag.max=40,type="correlation",
                        plot=TRUE)

hydroelectric_pacf <- Pacf(ts_energy_data[,3],lag.max=40,plot=TRUE)

#Combining plots
plot_grid(renewable_ts_plot,
          autoplot(renewable_acf),
          autoplot(renewable_pacf),
          nrow = 1, ncol = 3)

plot_grid(hydroelectric_ts_plot,
          autoplot(hydroelectric_acf),
          autoplot(hydroelectric_pacf),
          nrow = 1, ncol = 3)


```


### Q2
From the plot in Q1, do the series Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?

```{r}

#Total Renewable Energy Production
#This series does appear to have a trend -- the time series is increasing
#steadily as time goes on. Additionally, we note that the high correlation
#values in the ACF and the spikes in the PACF indicate a trend.

#Hydroelectric Power Consumption
#This series does appear to have a trend -- the time series is increasing
#steadily as time goes on. Additionally, we note that the high correlation
#values in the ACF and the spikes in the PACF indicate a trend.

```


### Q3
Use the *lm()* function to fit a linear trend to the three time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.

```{r}

nobs <- nrow(raw_energy_data)

t <- 1:nobs

renewable_linear_trend <- lm(raw_energy_data[,3]~t)
summary(renewable_linear_trend)

renewable_beta0 <- renewable_linear_trend$coefficients[1]
renewable_beta1 <- renewable_linear_trend$coefficients[2]

#For renewable energy, the slope is 0.70404 (Beta 0) and the intercept is 
#180.98940 (Beta 1). These values, in addition to the low p-value and
#higher R-squared, indicate a significant relationship between the renewable
#energy data and time. It is a moderately strong, positive relationship.

hydroelectric_linear_trend <- lm(raw_energy_data[,4]~t)
summary(hydroelectric_linear_trend)

hydroelectric_beta0 <- hydroelectric_linear_trend$coefficients[1]
hydroelectric_beta1 <- hydroelectric_linear_trend$coefficients[2]

#For hydroelectric power consumption, the slope is -0.009849 (Beta 0) and the
#intercept is 82.734747 (Beta 1). These values, in addition to the low p-value
#and lower R-squared, indicate a significant relationship between the
#hydroelectric data and time. It is a weak, negative relationship.

```

 
### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?

```{r}

renewable_y_detrend <- raw_energy_data[,3] - (renewable_beta0 + renewable_beta1*t)

renewable_df_detrend <- data.frame("date"=raw_energy_data[,1],
                                   "observed"=raw_energy_data[,2],
                                   "detrend"=renewable_y_detrend)

ggplot(renewable_df_detrend,aes(x=date))+
  geom_line(aes(y=observed),color="black")+
  geom_line(aes(y=detrend),color="blue")

hydroelectric_y_detrend <- raw_energy_data[,4] -(hydroelectric_beta0 + hydroelectric_beta1*t)

hydroelectric_df_detrend <- data.frame("date"=raw_energy_data[,1],
                                   "observed"=raw_energy_data[,3],
                                   "detrend"=hydroelectric_y_detrend)

ggplot(hydroelectric_df_detrend,aes(x=date))+
  geom_line(aes(y=observed),color="black")+
  geom_line(aes(y=detrend),color="blue")

#As compared to the plots from Q1, the detrended lines (shown in blue on both
#plots) have a mean of 0. These are significantly changed from the plots in Q1
#as the data has now been "de-trended". That being said, because a linear model
#was used to de-trend the data, we still see that there is some trend present.

```


### Q5
Plot ACF and PACF for the detrended series and compare with the plots from Q1. You may use plot_grid() again to get them side by side, but not mandatory. Did the plots change? How?

```{r}

renewable_acf_detrend <- Acf(renewable_df_detrend[,3],lag.max=40,
                             type="correlation",plot=TRUE)

renewable_pacf_detrend <- Pacf(renewable_df_detrend[,3],lag.max=40,plot=TRUE)

plot_grid(autoplot(renewable_acf),
          autoplot(renewable_pacf),
          autoplot(renewable_acf_detrend),
          autoplot(renewable_pacf_detrend))

hydroelectric_acf_detrend <- Acf(hydroelectric_df_detrend[,3],lag.max=40,
                                 type="correlation",plot=TRUE)

hydroelectric_pacf_detrend <- Pacf(hydroelectric_df_detrend[,3],lag.max=40,
                                   plot=TRUE)

plot_grid(autoplot(hydroelectric_acf),
          autoplot(hydroelectric_pacf),
          autoplot(hydroelectric_acf_detrend),
          autoplot(hydroelectric_pacf_detrend))

#For renewable energy, the plots changed slightly -- for the ACF, in the 
#detrended data, there are still very large correlation coefficients indicating
#that the trends were not totally removed. That said, the detrending had some
#effect as we can see a few spikes appearing at lag 12, lag 24, and lag 36 that
#indicate seasonality. The PACF, in the detrended data, has similar spikes but
#they are more pronounced as they "blossom" with less trend elsewhere in the
#data.

#For hydroelectric power consumption, the plots did not change and seem to
#appear largely the same. Based on the ACF and PACF, we see a similar
#sinusoidal pattern and spike pattern.

```


## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6
Just by looking at the time series and the acf plots, do the series seem to have a seasonal trend? No need to run any code to answer your question. Just type in you answer below.

```{r}

#For renewable energy, the time series does indicate a seasonal trend based on
#the wave-like patterns; the ACF, however, does not indicate a seasonal trend.

#For hydroelectric power consumption, the time series does indicate a seasonal 
#trend given the spikes that occur annually; the ACF does indicate a seasonal
#trend given the sinusoidal pattern that is evident.

```


### Q7
Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) the two time series. Ask R to print the summary of the regression. Interpret the regression output. From the results which series have a seasonal trend? Do the results match you answer to Q6?

```{r}

renewable_dummies <- seasonaldummy(ts_energy_data[,2])
renewable_seas_means_model <- lm(raw_energy_data[,3]~renewable_dummies)
summary(renewable_seas_means_model)

#For renewable energy, the high p-value and the low R-squared value indicate
#that the seasonal means model is not a good way to remove seasonality.

hydroelectric_dummies <- seasonaldummy(ts_energy_data[,3])
hydroelectric_seas_means_model <- lm(raw_energy_data[,4]~hydroelectric_dummies)
summary(hydroelectric_seas_means_model)

#For hydroelectric, the low p-value and the higher R-squared value indicate
#that the seasonal means model is a good fit for this data.

#In the case of renewable energy, it's not that there is no seasonality, it's
#just that a seasonal means model is not a good representation of the data. For
#hydroelectric, the results do match the answer to Q6.


```


### Q8
Use the regression coefficients from Q7 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?

```{r}

#Not running this for renewable energy since there was no seasonal trend
#identified in the seasonal means model

#Hydroelectric
#Look at the regression coefficient. These will be the values of Beta
#Store regression coefficients
hydroelectric_beta_int <- hydroelectric_seas_means_model$coefficients[1]
hydroelectric_beta_coeff <- hydroelectric_seas_means_model$coefficients[2:12]

#compute seasonal component
hydroelectric_seas_comp <- array(0,nobs)
for(i in 1:nobs){
  hydroelectric_seas_comp[i] <- (hydroelectric_beta_int+hydroelectric_beta_coeff
                                 %*% hydroelectric_dummies[i,])
}

#Understanding what we did
ggplot(raw_energy_data, aes(x=Month, y=raw_energy_data[,4])) +
            geom_line(color="blue") +
            ylab(paste0(colnames(raw_energy_data)[4],sep="")) +
            geom_line(aes(y=hydroelectric_seas_comp), col="red")

#Removing seasonal component
deseason_hydroelectric_data <- raw_energy_data[,4]-hydroelectric_seas_comp

#Understanding what we did
ggplot(raw_energy_data, aes(x=Month, y=raw_energy_data[,4])) +
            geom_line(color="blue") +
            ylab(paste0(colnames(raw_energy_data)[4],sep="")) +
            geom_line(aes(y=deseason_hydroelectric_data), col="green")

#Compared to Q1, the plots move the mean to 0 and the regular, wave-like
#pattern is gone. The seasonal component has been removed.

```


### Q9
Plot ACF and PACF for the deseason series and compare with the plots from Q1. You may use plot_grid() again to get them side by side, but not mandatory. Did the plots change? How?

```{r}

#Not running this for renewable energy since there was no seasonal trend
#identified in the seasonal means model

hydroelectric_acf_deseason <- Acf(deseason_hydroelectric_data,lag.max=40,
                                  type="correlation", plot=TRUE)

hydroelectric_pacf_deseason <- Pacf(deseason_hydroelectric_data,lag.max=40,
                                    plot=TRUE)

plot_grid(autoplot(hydroelectric_acf),
          autoplot(hydroelectric_pacf),
          autoplot(hydroelectric_acf_deseason),
          autoplot(hydroelectric_pacf_deseason))

#For hydroelectric power consumption, the plots did change. Notably, the ACF
#moves from a sinusoidal pattern to a decreasing trend. Around lag 20, the
#correlation becomes insignificant. The PACF is telling us that we only need to
#add t - 1 to our data set. In other words, significant de-trending has
#occurred here.

```



